---
title: 'EDS241: Final'
author: "Joe DeCesaro"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output:
  pdf_document:
    toc: false
    number_sections: yes
header-includes:
- \setlength{\parindent}{1em}
- \usepackage{float}
---

``` {r setup, echo=FALSE, message=FALSE, warning = FALSE}
# set default chunk options
knitr::opts_chunk$set(fig.width = 4, fig.height = 3, echo = TRUE, message = FALSE, warning = FALSE)

# load packages
packages = c("stargazer", "here", "tidyverse", "janitor", 
           "cowplot", "tinytex", "datasets", "estimatr", "tibble", "car", 
           "readxl", "huxtable", "AER", "knitr")

for (i in packages) {
  if (require(i, character.only = TRUE) == FALSE) {
    install.packages(i,repos = 'http://cran.us.r-project.org')
  }
  else {
    require(i, character.only = TRUE)
  }
}

options(scipen = 5) # not scientific notation
```

# Take Home Final

\noindent The question for this take-home final exam asks you to examine the impact of the opening of a garbage incinerator on housing values in North Andover, MA. The data for the exercise are a subset of the data in the paper: K.A. Kiel and K.T. McClain (1995): “House Prices During Siting Decision Stages: The Case of an Incinerator from Rumor Through Operation,” Journal of Environmental Economics and Management 28, 241-255.

## Background:
\noindent The construction of a new garbage incinerator in North Andover in the early 1980s was controversial due to the increases in ambient pollution that it would create. Rumors of the incinerator began after 1978. The construction started in 1981, and the incinerator began operating in 1985. In Economics, land market theory suggests that local amenities are capitalized in housing values, and predicts that the prices of houses located near the incinerator would fall compared to the price of houses located further away from the incinerator. By 1981, you can assume that all market participants had full information on the upcoming garbage incinerator, so that housing values had capitalized the upcoming arrival of the incinerator.

## Data:
\noindent The authors of the paper collected data on prices of houses that sold in 1978 (before the upcoming construction of the incinerator was public knowledge) and in 1981 (after the construction had started). The key variables for the analysis are: 
- rprice (inflation-adjusted sales price of house)
- nearinc (=1 if house located near the incinerator, =0 otherwise)
- age (age of the house)
- land (square footage of the lot)
- area (square footage of the house)
- rooms (number of rooms in the house)
- year indicator (1978 or 1981). 

These variables are contained in the CSV file KM_EDS241.csv.

### Read in the data
```{r}
house_data <-read.csv(here("KM.csv"))
```

## (a) Using the data for 1981, estimate a simple OLS regression of real house values on the indicator for being located near the incinerator in 1981. What is the house value “penalty” for houses located near the incinerator? Does this estimated coefficient correspond to the ‘causal’ effect of the incinerator (and the negative amenities that come with it) on housing values? Explain why or why not.

```{r}
house_data_1981 <- house_data %>% 
  filter(year == 1981) # filter to just get 1981

model1 <- lm_robust(rprice ~ nearinc, # effect of being near incinerator on prices in 1981
                    data = house_data_1981) 
summary(model1)
```

\noindent The house "penalty" for houses near the incinerator is an average decrease in value of `r -1*round(model1$coefficients[[2]], 2)` dollars compared to houses not near the incinerator. The estimated coefficient corresponds with the 'causal' effect of the incinerator as the p-value deems it a statistically significant difference at the 0.001 level. However, the incinerator could have been placed in an area that already had lower housing prices generally so this could be omitted variables bias. This is explored in the next question.

## (b) Using the data for 1978, provide some evidence the location choice of the incinerator was not “random”, but rather selected on the basis of house values and characteristics. [Hint: in the 1978 sample, are house values and characteristics balanced by nearinc status?]
```{r}
# lets check if the prices were lower in these areas already
house_data_1978 <- house_data %>% 
  filter(year == 1978)

# house prices that would be near/away incinerator in 1978
model2.1 <-  lm_robust(formula = rprice ~ nearinc, 
                       data = house_data_1978)

model2.1_sum <- summary(model2.1)
model2.1_sum
```

\noindent Based on the housing information from 1978, the incinerator was placed near houses that, on average, were sold for `r -1*round(model2.1$coefficients[[2]], 2)` adjusted dollars less than houses not near the incinerator placement. The p-value for the difference in houses near the incinerator is `r round(model2.1_sum$coefficients[2,4], 5)` and therefore statistically significant at the 0.01 level. This indicates that placement of the incinerator was not random but near homes that were of less value already, lets look at another variable to be sure.

```{r}
# did houses near the incinerator have more, less, or the same land?
model2.2 <-  lm_robust(formula = land ~ nearinc, 
                       data = house_data_1978)

model2.2_sum <- summary(model2.2)
model2.2_sum
```

\noindent Based on the housing information from 1978, the incinerator was placed near houses that, on average, had `r -1*round(model2.2$coefficients[[2]], 2)` less of lot square footage than houses not near the incinerator placement. The p-value for the difference in houses near the incinerator is `r round(model2.2_sum$coefficients[2,4], 5)` and therefore statistically significant at the 0.001 level. This confirms that placement of the incinerator was not random.

## (c) Based on the observed differences in (b), explain why the estimate in (a) is likely to be biased downward (i.e., overstate the negative effect of the incinerator on housing values).

\noindent The difference in housing prices near the incinerator in 1982 is likely affected by the fact that they were on average worth less and had smaller lot sizes than those homes further away from the incinerator. The affect on price we are seeing in (a) is likely overstated from omitting variables from the regression and they should be controlled for.

## (d) Use a difference-in-differences (DD) estimator to estimate the causal effect of the incinerator on housing values without controlling for house and lot characteristics. Interpret the magnitude and sign of the estimated DD coefficient.

```{r}
 house_data <-  house_data %>% 
  # make binary variable for years
  mutate(post1981 = case_when(year == 1978 ~ 0,
                                      year == 1981 ~ 1),
         # make treatment dummy binary variable for houses in 1981 that are near the inc
         treatment_dummy = nearinc*post1981) 

# make the DD model for different binary possibilities
model3 <- lm_robust(formula = rprice ~ post1981 + nearinc + treatment_dummy,
                    data = house_data)

# to be able to get the confidence intervals by calling them
model3_table <- tidy(model3) %>%  
  select(term, estimate, std.error, p.value, conf.low, conf.high)

model3_sum <- summary(model3)
model3_sum
```

\noindent The estimated DD causal effect of the incinerator on housing values without controlling for other characteristics is an adjusted sales price difference of `r round(model3$coefficients[[4]], 2)`. This value represents the difference between adjusted sales price of houses compared to those that are not near the incinerator and sold in 1981 (control) AND the difference compared to houses sold in 1978 and would be near the incinerator (treated pre-policy intervention). The negative sign means that the houses in the area that were near the incinerator (treated) had a lower price sales than those not treated and sold in 1978. It should be noted that the p-value is `r round(model3_sum$coefficients[4,4], 3)` and therefore not significant.

## (e) Report the 95% confidence interval for the estimate of the causal effect on the incinerator in (d).
The 95% confidence interval for the estimate of the causal effect on the incinerator is[`r round(model3_table$conf.low[4], 2)`, `r round(model3_table$conf.high[4], 2)`]. The 95% confidence interval crosses 0 so the DD estimate is not statistically different from zero.

## (f) How does your answer in (d) changes when you control for house and lot characteristics? Test the hypothesis that the coefficients on the house and lot characteristics are all jointly equal to 0.

```{r}
# make the DD model for different binary possibilities and control for housing characteristics
model4 <- lm_robust(formula = rprice ~ post1981 + nearinc + treatment_dummy + 
                      # control variables
                      age + rooms + area + land, 
                    data = house_data)

model4_table <- tidy(model4) %>%
  select(term, estimate, std.error, p.value, conf.low, conf.high)

summary(model4)
```

\noindent The estimated DD causal effect of the incinerator on housing values while controlling for other characteristics is an adjusted sales price difference of `r round(model4$coefficients[[4]], 2)`. The negative sign means that the houses in the area that were near the incinerator (treated) had a lower price sales than those not treated and sold in 1978. It should be noted that the confidence interval does cross 0 so the DD estimate is not statistically different from zero.

```{r}
# linear hypothesis test
hyp_test <- linearHypothesis(model4, 
                             c("age = 0", "rooms = 0", "area = 0", "land = 0"),
                             white.adjust = "hc2")
hyp_test
```

\noindent The proper F-statistic is `r round(hyp_test$Chisq[[2]], 0)` which is greater than 10, therefore these are not a weak instrument. The p-value of the linear hypothesis test is statistically significant we can reject the null hypothesis that the coefficients on the housing characteristics are all jointly equal to 0.

## (g) Using the results from the DD regression in (f), calculate by how much did real housing values change on average between 1978 and 1981. (for the control group per Slack)

\noindent The real housing values change on average between 1978 and 1981 for the control group when controlling for other housing characteristics is `r round(model4$coefficients[[2]], 2)`.

## (h) Explain (in words) what is the key assumption underlying the causal interpretation of the DD estimator in the context of the incinerator construction in North Andover.




